Reasonably Programmable Literal Notation

### 2018-03-09 icfp18/syntax-icfp18.tex
rewrite of sec 2.1 -- client perspective

Although requests for specialized literal notation are easy to dismiss as superficial, the reality is that literal notation, or the absence thereof, can have a substantial influence on software quality. For example, \citet{Bravenboer:2007:PIA:1289971.1289975} finds that literal notation for structured encodings of queries, like the SQL-based query literals now found in many languages \cite{meijer2006linq}, reduce the temptation to use string encodings of queries and therefore reduce the risk of catastrophic string injection attacks~\cite{owasp2017}. More generally, evidence suggests that programmers frequently resort to ``stringly-typed programming'', i.e. they choose strings instead of composite data structures largely for reasons of notational convenience. In particular, \citet{TSLs} sampled strings from open source projects and found that at least 15\% of them could be parsed by some readily apparent type-specific grammar, e.g. for URLs, paths, regular expressions and many others. 
Literal notation, with support for splicing, would decrease the syntactic cost of composite encodings, which are more amenable to programmatic manipulation and compositional reasoning. %Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.

Of course, it would not be scalable to ask general-purpose language designers to build in support for all known notations \emph{a priori}. Instead, there has been persistent interest in mechanisms that allow library providers to define new literal notation on their own. For example, direct grammar extension systems like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, macro-based systems like Template Haskell~\cite{SheardPeytonJones:Haskell-02,mainland2007s}, the system of \emph{type-specific languages} (TSLs) described by \citet{TSLs}, and other systems that we will discuss below can all be used to define new literal notation (and, in some cases, other forms of new notation, such as new infix expression forms, control flow operations or type declaration forms, which we leave beyond the scope of this paper).

\paragraph{Problem} The problem that specifically motivates this paper is that these existing systems  make it difficult or impossible to reason abstractly about such fundamental issues as types and variable binding. Instead, programmers and editor services can only reason transparently, i.e. by inspecting the underlying expansion or the implementation details of the collection of extensions responsible for producing the expansion.% In some existing systems, even determining which syntax extension is responsible for a literal form is difficult.
%, i.e. when developing large programs ``consisting of many small programs o(modules), possibly written by different people''
\begin{figure}[t!]
\begin{subfigure}[t]{0.48\textwidth}
\fcolorbox{gray!20}{gray!20}{\makebox[0.97\textwidth][l]{\textsc{existing grammar extension systems}}}
\vspace{-12px}
\begin{lstlisting}[xleftmargin=-2pt, morekeywords={EXTEND}]
  EXTEND /* loaded by, e.g., camlp4 */
    expr:
      |  "`(" q = kquery ")`" -> q
    kquery: 
      | /* ...K query grammar... */
  /* ...more extensions defined... */
  let w = compute_w();
  let x = compute_x(w);
  let y = `((!R)@&{&/x!/:2_!x}'!R)`;
\end{lstlisting}
\vspace{-5px}
\caption{It is difficult to reason abstractly about programs that use unfamiliar grammar extensions (see the six reasoning criteria in the text).}
\label{fig:K-dialect}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.51\textwidth}
\fcolorbox{gray!20}{gray!20}{\makebox[\textwidth][l]{\textsc{\textbf{this paper}: typed literal macros}}}
\vspace{-12px}
\begin{lstlisting}[xleftmargin=-2pt]
  notation $kq at KQuery.t 
    lexer  KQueryLexer
    parser KQueryParser.start
    expansions require
      module KQuery as KQuery;
  /* ...more notations defined... */
  let w = compute_w();
  let x = compute_x(w);
  let y = $kq `(SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!REURL)`;
\end{lstlisting}
\vspace{-5px}
\caption{TLMs make examples like these more reasonable by leaving the base grammar fixed and making the type and binding structure and the segmentation explicit.}
\label{fig:K-tsm-example}
\end{subfigure}
\vspace{4px}
\caption{Two of the possible ways to introduce literal notation for encodings of K queries}
\vspace{-10px}
\end{figure}

Consider, for example, the perspective of a programmer attempting to comprehend the program text in Figure \ref{fig:K-dialect}, which is written in a dialect of OCaml's surface syntax called Reason \cite{reason-what} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system --- Lines 1-5 show the grammar extension syntax from Camlp4 \cite{ocaml-manual}, but Sugar*/SugarHaskell is  similar \cite{erdweg2011sugarj,erdweg2013framework,erdweg2012layout}. 
Line 9 uses one of the active syntax extensions to construct an encoding of a query in the niche database query language K, using its intentionally terse notation~\cite{Whitney:2001:LOR:376284.375783}. The problem is that a programmer looking at the program as presented, and unfamiliar with (i.e. holding abstract) the details elided on Lines 5-6, cannot easily answer questions like the following:

\begin{enumerate}[leftmargin=15px]
\item \textbf{Responsibility}: Which syntax extension determined the expansion of the literal on Line~9? Might activating a new extension generate a conflicting expansion for the same literal?
\item \textbf{Expansion Typing}: What type does \li{y} have?
\item \textbf{Context Dependence}: Which bindings does the expansion of Line~9 invisibly depend on? If I shadow or remove a module or other binding, could that break or change the meaning of Line~9 because its expansion depends invisibly on the original binding?
\item \textbf{Segmentation}: Are the characters \li{x}, \li{R} and \li{2} on Line~9 parsed as spliced terms, meaning that they appear directly in the underlying expansion, or are they parsed in some other way peculiar to this literal notation (e.g. as operators in the K query language)?
\item \textbf{Segment Typing}: What type is each spliced term expected to have? How can I infer a type for a variable that appears in a spliced term without looking at where it ends up in the expansion?
\item \textbf{Capture}: If \li{x} is in fact a spliced term, does it refer to the binding of \li{x} on Line 5, or might it capture an invisible binding of the same identifier in the expansion of Line 9?
\end{enumerate}

@@

Given these issues, we concluded that direct grammar extension systems were not ideally suited for integration into Reason. We also evaluated various approaches that are based not on direct grammar extension but on term rewriting over a fixed grammar. We give a full account of this evaluation in Sec.~\ref{sec:existing-approaches}. Briefly, we found that:
\begin{itemize}[leftmargin=15px]
\item Unhygienic approaches like OCaml's preprocessor extension point (PPX) rewriters \cite{ocaml-manual} and Template Haskell \cite{SheardPeytonJones:Haskell-02,mainland2007s} allow us to define new literal notation with support for splicing by repurposing existing string literal forms. They also partially or completely solve the problem of reasoning about \textbf{Responsibility} by using explicit annotations on terms that are being rewritten. However, they do not satisfy the remaining five reasoning criteria.
\item Hygienic staging macro systems, like MetaOCaml and related systems \cite{DBLP:conf/flops/Kiselyov14,ganz2001macros,Sheard:1999:UMS} do not give the expander access to the syntax trees of arguments, i.e. the rewriting is parametric in the arguments, so they are not suitable for the problem of defining new literal notation.  
\item Hygienic term rewriting macro systems, like those in various Lisp-family languages \cite{DBLP:conf/esop/HermanW08,Herman10:Theory,Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92,mccarthy1978history,Flatt:2012:CLR:2063176.2063195} (e.g. Racket \cite{Flatt:2012:CLR:2063176.2063195}) and Scala \cite{ScalaMacros2013}, do not allow us to repurpose string literal forms to define literal forms at composite types because the hygiene discipline does not allow us to splice terms out of literal bodies via parsing. Only existing composite forms can therefore be repurposed.
\item \emph{Type-specific languages} (TSLs) \cite{TSLs} come closer to our goals in that they explicitly support splicing terms out of literal bodies, but the mechanism falls short with regard to the six reasoning principles just discussed in ways that we will return to in Sec. \ref{sec:existing-approaches}. In any case, this approach was designed for simple nominally-typed languages and relies critically on a particular local type inference scheme. It is not suitable for a language with an ML-like semantics, meaning a language with support for structural types (like tuple and function types in ML), parameterized types, abstract types, modules, pattern matching and non-local type inference.
\end{itemize}

@@

\paragraph{Brief Overview} For a brief overview of our proposed mechanism, consider Figure \ref{fig:K-tsm-example}. On Lines 1-5, we define a TLM named \li{$kq} that provides K query literal notation. On Line 9, we apply this TLM to express the example from Figure \ref{fig:K-dialect}. We can reason abstractly about this program as follows.
\begin{enumerate}[leftmargin=12pt]
\item \textbf{Responsibility}: The lexer and parser specified by the applied TLM on Lines 2-3 are exclusively responsible for lexing, parsing and expanding the body of the generalized literal form, i.e. the characters between \li{`(} and \li{)`}. We will give more details on generalized literal forms and on constructing a TLM lexer and parser in the next section. For now, let us simply reiterate that our design goal is to provide a system where the programmer does not normally need to look up the definitions of \li{KQueryLexer} and \li{KQueryParser} to reason about types and binding. % To briefly summarize, our system uses Menhir, which is the most popular parser generator in the OCaml ecosystem and integrated into its standard build tools. %the context-free grammar of the language is fixed, so the TLM provider can modularly establish that the notation that the TLM implements is unambiguous (even in situations where a spliced term itself applies another TLM, which we will show in a later example).
\item \textbf{Expansion Typing}: The type annotation on the definition of \li{$kq} (Line 1) determines the type that the expansion must have, here \li{KQuery.t}.
\item \textbf{Context Dependence}: Lines 4-5 specify that expansions generated by \li{$kq} are allowed to use the module \li{KQuery}, and no others. The system ensures that this dependency is  bound as specified even if the identifier \li{KQuery} has been shadowed at the application site. This completely relieves clients from needing to consider expansion-internal dependencies when naming variables. %For example, if the client renames \li{w}, there is not even a remote possibility that this could break or change the meaning of Line 9.
\item \textbf{Segmentation}: The intermediate output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms, which appear directly in the expansion, from segments that are parsed in some other way by the TLM. The segmentation is all that needs to be communicated to language services downstream of the expander, e.g. editors and pretty printers, which can communicate this information to the programmer using secondary notation, e.g. colors in this document. So by examining Fig. \ref{fig:K-tsm-example}, the programmer knows that the two instances of \li{x} on Line 3 are parsed as spliced expressions (because they are black), whereas the \li{R}'s must be parsed in some other way, e.g. as operators of the K language (because they are green).
\item \textbf{Segment Typing}: Each spliced segment in the inferred segmentation also has  a type annotation. This, together with the context independence property, ensures that type inference at the TLM application site can be performed (by editor services or in the programmer's mind) abstractly, i.e. by reference only to the type annotations on the spliced segments, not the full expansion.  %This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired. %TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kq}.
\item \textbf{Capture}: Splicing is guaranteed to be capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2. It cannot have captured a coincidental binding of \li{x} in the expansion. % We will say more about capture later on.
\end{enumerate}

@@

\paragraph{Paper Outline} 
% \begin{itemize}
% \item 
Sec.~\ref{sec:setsms} introduces expression TLMs with a substantially more detailed case study. 
% \item 
Sec.~\ref{sec:sptsms} then introduces {pattern TLMs} and describes the special reasoning conditions in pattern position. %The hygiene condition for pattern TLMs is interesting.
% \item 
Sec.~\ref{sec:ptsms} introduces the more general {parametric TLMs}, i.e. TLMs that can take type and module parameters, which allow us to define literal notation over a type- or module-parameterized family of types. This section also reveals that the \li{expansions require} clause can be understood as parameterization followed immediately by partial parameter application. 
Having introduced the basic machinery, we continue in Sec.~\ref{sec:more-examples} with examples that further demonstrate the expressive power of this approach and various parser implementation strategies that notation providers can consider. Notably, this section gives our take on the example from  Fig.~\ref{fig:urweb} of Ur/Web-style HTML literals as well as a TLM that implements quasiquotation for Reason language terms, which is useful for writing other TLMs. 
Next, Sec.~\ref{sec:setsms-formally} defines a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above in their simplest form. 
Sec.~\ref{sec:ptlms-formally} adds type functions and an ML-style module system to this calculus, and gives a more general variant of the reasoning principles theorem. 
% \item 
Sec.~\ref{sec:existing-approaches} compares TLMs to related work, guided by the rubric of  reasoning principles just discussed. 
% \item 
Sec.~\ref{sec:discussion} concludes with a summary of the contributions of this paper and a discussion of limitations and future work.
Certain technical details, proofs and extensions to the calculi presented in this paper are given in the supplement. % Also available is an additional series of case studies: notation for an encoding of regular expressions as an abstract data type, implemented using TLMs for quasiquotation and grammar-based parser generators. %are also available in the supplement. Sec.~\ref{sec:static-eval} considers the topic of term evaluation during the  expansion phase, i.e. \emph{static evaluation}, in  more detail. Sec \ref{sec:static-eval} also gives examples of TLMs that are useful for defining other TLMs, e.g. TLMs that implement parser generators and quasiquotation.

@@

Consider the recursive datatype \li{Regex.t} defined by Fig.~\ref{fig:Regex-module-def}, which encodes {regular expressions} (regexes) into Reason \cite{Thompson:1968:PTR:363347.363387}. Regexes are  common in, for example, bioinformatics, where they are used to express patterns in DNA sequences. We can construct a regex that matches the strings \li{"A"}, \li{"T"}, \li{"G"} or \li{"C"}, which represent the four bases in DNA, as follows:
\begin{lstlisting}[numbers=none]
  let any_base = Regex.(Or(Str "A", Or(Str "T", Or(Str "G", Or(Str "C")))))
\end{lstlisting}
In Reason, the notation \li{Regex.(}$e$\li{)} locally opens the module \li{Regex} within the expression $e$, so we did not need to qualify each application of \li{Regex.Or} and \li{Regex.Str} above. Even with this shorthand, however, constructing regexes in this way is syntactically costly. Instead, we would like to have the option to use the common POSIX-style notation  \cite{STD95954} when constructing values of type \li{Regex.t}, including values constructed compositionally from previously constructed regexes and strings. 
We solve this problem in Fig. \ref{fig:regex-tlm-def} by defining a TLM named \li{$regex} (pronounced ``lit regex'') that supports POSIX-style regex notation extended with splice forms for regexes and strings. 
% One approach would be to simply define a function \li{parse_rx : string -> Regex.t} that parses a string at run-time to produce a regex. This approach works for simple expressions like the one above -- we can write \li{parse_rx "A|T|G|C"}. However, this approach breaks down when one needs to construct regexes compositionally. For example, the DNA pattern that the BisI restriction enzyme recognizes is \li{AT}$x$\li{GC}, where $x$ is any of the four DNA bases. However, one cannot pass \li{any_base} to the string concatenation operator, because \li{any_base} is already of type \li{Regex.t}. These kinds of situations cause programmers to engage in \emph{stringly-typed programming}, as discussed in Sec.~\ref{sec:intro} -- they simply use strings to encode regexes throughout their application, rather than structured encodings like \li{Regex.t}. This can be catastrophic if the programmer is not extremely careful with string concatenation because strings that happen to contain special characters recognized by the POSIX notation can be misinterpreted because they are not distinguished in any way from strings that intentionally use POSIX notation. When some of these strings come from untrusted sources, e.g. web forms as is common in bioinformatics applications, the result can be a breach in privacy.

% Moreover, there already exists a broadly adopted shorthand notational standard -- the POSIX standard for regular expressions \cite{STD95954}. For the purposes of this section, we will stick to the subset of the POSIX standard that applies to the encoding in Fig.~\ref{fig:Regex-module-def}, but there is nothing that would fundamentally prevent us from implementing the full POSIX standard using TLMs.
\begin{figure}[t]
\begin{subfigure}[t]{0.45\textwidth}
\vspace{-1px}
\begin{lstlisting}[mathescape=~]
  module Regex = {
    type t = AnyChar | Str(string)
           | Seq(t, t) | Or(t, t) 
           | Star(t);
  };
\end{lstlisting}
\vspace{-5px}
\caption{The \li{Regex} module, which defines the recursive datatype \li{Regex.t}.}
\label{fig:Regex-module-def}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.53\textwidth}
\vspace{-1px}
\begin{lstlisting}[mathescape=|]
  notation $regex at Regex.t
    lexer   RegexLexer
    parser  RegexParser.start
    expansions require
      module Regex as Regex;
\end{lstlisting}
\vspace{-5px}
\caption{The definition of the \li{$regex} TLM. Figure \ref{fig:lexer-and-parser} defines \li{RegexLexer} and \li{RegexParser}.}
\label{fig:regex-tlm-def}
\end{subfigure}

>>>>

Although requests for specialized literal notation are easy to dismiss as superficial, the reality is that literal notation, or the absence thereof, can have a substantial influence on software quality. For example, \citet{Bravenboer:2007:PIA:1289971.1289975} finds that literal notation for structured encodings of queries, like the SQL-based query literals now found in many languages \cite{meijer2006linq}, reduce the temptation to use string encodings of queries and therefore reduce the risk of catastrophic string injection attacks~\cite{owasp2017}. More generally, evidence suggests that programmers frequently resort to ``stringly-typed programming'', i.e. they choose strings instead of composite data structures, largely for reasons of notational convenience. In particular, \citet{TSLs} sampled strings from open source projects and found that at least 15\% of them could be parsed by some readily apparent type-specific grammar, e.g. for URLs, paths, regular expressions and many others. 
Literal notation, with support for splicing, would decrease the syntactic cost of composite encodings, which are more amenable to programmatic manipulation and compositional reasoning. %Literal forms would  preferable to string encodings because they support compositional construction and decomposition by pattern matching via splicing.

Of course, it would not scale to ask general-purpose language designers to build in support for all known notations \emph{a priori}. Instead, there has been persistent interest in mechanisms that allow library providers to define new literal notation on their own. For example, direct grammar extension systems like Camlp4 \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, term rewriting systems like Template Haskell~\cite{SheardPeytonJones:Haskell-02,mainland2007s}, the system of \emph{type-specific languages} (TSLs) described by \citet{TSLs}, and other systems that we will discuss below can all be used to define new literal notation (and, in some cases, other forms of new notation, such as new infix expression forms, control flow operations or type declaration forms, which we leave beyond the scope of this paper).

\paragraph{Problem} The problem that specifically motivates this paper is that these existing systems  make it difficult or impossible to reason abstractly about such fundamental issues as types and variable binding when presented with a program using user-defined literal forms. Instead, programmers and editor services can only reason transparently, i.e. by inspecting the underlying expansion or the implementation details of the collection of extensions responsible for producing the expansion.% In some existing systems, even determining which syntax extension is responsible for a literal form is difficult.
%, i.e. when developing large programs ``consisting of many small programs o(modules), possibly written by different people''
\begin{figure}[t!]
\begin{subfigure}[t]{0.48\textwidth}
\fcolorbox{gray!20}{gray!20}{\makebox[0.97\textwidth][l]{\textsc{existing grammar extension systems}}}
\vspace{-12px}
\begin{lstlisting}[xleftmargin=-2pt, morekeywords={EXTEND}]
  EXTEND /* loaded by, e.g., camlp4 */
    expr:
      |  "`(" q = kquery ")`" -> q
    kquery: 
      | /* ...K query grammar... */
  /* ...more extensions defined... */
  let w = compute_w();
  let x = compute_x(w);
  let y = `((!R)@&{&/x!/:2_!x}'!R)`;
\end{lstlisting}
\vspace{-5px}
\caption{It is difficult to reason abstractly about programs that use unfamiliar grammar extensions (see the six reasoning criteria in the text).}
\label{fig:K-dialect}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.51\textwidth}
\fcolorbox{gray!20}{gray!20}{\makebox[\textwidth][l]{\textsc{\textbf{this paper}: typed literal macros}}}
\vspace{-12px}
\begin{lstlisting}[xleftmargin=-2pt]
  notation $kq at KQuery.t 
    lexer  KQueryLexer
    parser KQueryParser.start
    expansions require
      module KQuery as KQuery;
  /* ...more notations defined... */
  let w = compute_w();
  let x = compute_x(w);
  let y = $kq `(SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!REURL)`;
\end{lstlisting}
\vspace{-5px}
\caption{TLMs make examples like these more reasonable by leaving the base grammar fixed and making the type and binding structure and the segmentation explicit.}
\label{fig:K-tsm-example}
\end{subfigure}
\vspace{4px}
\caption{Two of the possible ways to introduce literal notation for encodings of K queries}
\vspace{-10px}
\end{figure}

Consider, for instance, the perspective of a programmer trying to comprehend the program text in Fig. \ref{fig:K-dialect}, which is written in a  popular dialect of OCaml's surface syntax called Reason \cite{reason-what}\todo{up-to-date?}{} that has, hypothetically, been extended with some number of new literal forms by a grammar extension system --- Lines 1-5 show the grammar extension syntax from Camlp4 \cite{ocaml-manual}, but Sugar*/SugarHaskell is  similar \cite{erdweg2011sugarj,erdweg2013framework,erdweg2012layout}. 
Line 9 uses one of these active syntax extensions to construct an encoding of a query in the niche database query language K, using its intentionally terse notation~\cite{Whitney:2001:LOR:376284.375783}. The problem is that a programmer looking at the program as presented, and unfamiliar with (i.e. holding abstract) the details elided on Lines 5-6, cannot easily answer questions like the following:

\begin{enumerate}[leftmargin=15px]
\item \textbf{Responsibility}: Which syntax extension determined the expansion of the literal on Line~9? Might activating a new extension generate a conflicting expansion for the same literal?
\item \textbf{Expansion Typing}: What type does \li{y} have?
\item \textbf{Context Dependence}: Which bindings does the expansion of Line~9 invisibly depend on? If I shadow or remove a module or other binding, could that break or change the meaning of Line~9 because its expansion depends invisibly on the original binding?
\item \textbf{Segmentation}: Are the characters \li{x}, \li{R} and \li{2} on Line~9 parsed as spliced terms, meaning that they appear directly in the underlying expansion, or are they parsed in some other way peculiar to this literal notation, e.g. as operators in the K query language?
\item \textbf{Segment Typing}: What type is each spliced term expected to have? How can I infer a type for a variable that appears in a spliced term without looking at where it ends up in the expansion?
\item \textbf{Capture}: If \li{x} is in fact a spliced term, does it refer to the binding of \li{x} on Line 5, or might it capture an invisible binding of the same identifier in the expansion of Line 9?
\end{enumerate}

@@

Given these issues, we concluded that direct grammar extension systems were not ideally suited for integration into the Reason platform, which seeks to develop a clear and modern surface syntax for the OCaml programming language. We also evaluated various approaches that are based not on direct grammar extension but on term rewriting over a fixed grammar. We give a full account of this evaluation in Sec.~\ref{sec:existing-approaches}, but briefly, we found that:
\begin{itemize}[leftmargin=15px]
\item Unhygienic approaches like OCaml's preprocessor extension point (PPX) rewriters \cite{ocaml-manual} and Template Haskell \cite{SheardPeytonJones:Haskell-02,mainland2007s} allow us to define new literal notation with support for splicing by repurposing existing string literal forms. They also partially or completely solve the problem of reasoning about \textbf{Responsibility} by using explicit annotations on terms that are being rewritten. However, they do not satisfy the remaining five reasoning criteria.
\item Hygienic staging macro systems, like MetaOCaml and related systems \cite{DBLP:conf/flops/Kiselyov14,ganz2001macros,Sheard:1999:UMS} do not give the expander access to the syntax trees of arguments, i.e. the rewriting is parametric in the arguments, so they are not suitable for the problem of defining new literal notation.  
\item Hygienic term rewriting macro systems, like those in various Lisp-family languages \cite{DBLP:conf/esop/HermanW08,Herman10:Theory,Kohlbecker86a,DBLP:conf/popl/Adams15,DBLP:conf/popl/ClingerR91,DBLP:journals/lisp/DybvigHB92,mccarthy1978history,Flatt:2012:CLR:2063176.2063195}, e.g. Racket \cite{Flatt:2012:CLR:2063176.2063195}, as well as Scala \cite{ScalaMacros2013}, do not allow us to repurpose string literal forms to define composite literal forms because the hygiene discipline does not allow us to splice terms out of literal bodies via parsing (the system cannot identify these terms as sub-terms of the macro arguments \cite{DBLP:conf/esop/HermanW08,Herman10:Theory}).% Only existing composite forms can therefore be repurposed.
\item \emph{Type-specific languages} (TSLs) \cite{TSLs} come closer to our goals in that they explicitly support splicing terms out of literal bodies, but the mechanism falls short with regard to the six reasoning principles just discussed in ways that we will return to in Sec. \ref{sec:existing-approaches}. In any case, this approach was designed for simple nominally-typed languages and relies critically on a particular local type inference scheme. It is not suitable for a language with an ML-like semantics, meaning a language with support for structural types (like tuple and function types in ML), parameterized type families, pattern matching and non-local type inference.
\end{itemize}

@@

\paragraph{Brief Overview} For a brief overview of our proposed mechanism, consider Fig.~\ref{fig:K-tsm-example}. On Lines~1-5, we define a TLM named \li{$kq} that provides K query literal notation. On Line 9, we apply this TLM to express the example from Figure \ref{fig:K-dialect}. We can reason abstractly about this program as follows.
\begin{enumerate}[leftmargin=12pt]
\item \textbf{Responsibility}: The lexer and parser specified by the applied TLM on Lines 2-3 are together exclusively responsible for lexing, parsing and expanding the body of the generalized literal form, i.e. the characters between \li{`(} and \li{)`}. We will give more details on generalized literal forms and on constructing a TLM lexer and parser in the next section. For now, let us simply reiterate that our design goal is to provide a system where the programmer does not normally need to look up the definitions of \li{KQueryLexer} and \li{KQueryParser} to reason about types and binding. % To briefly summarize, our system uses Menhir, which is the most popular parser generator in the OCaml ecosystem and integrated into its standard build tools. %the context-free grammar of the language is fixed, so the TLM provider can modularly establish that the notation that the TLM implements is unambiguous (even in situations where a spliced term itself applies another TLM, which we will show in a later example).
\item \textbf{Expansion Typing}: The type annotation on the definition of \li{$kq} (Line 1) determines the type that the expansion must have, here \li{KQuery.t}.
\item \textbf{Context Dependence}: Lines 4-5 specify that expansions generated by \li{$kq} are allowed to use the module \li{KQuery}, and no others. The system ensures that this dependency is  bound as specified even if the variable \li{KQuery} has been shadowed at the application site. This completely relieves clients from needing to consider expansion-internal dependencies when naming variables. %For example, if the client renames \li{w}, there is not even a remote possibility that this could break or change the meaning of Line 9.
\item \textbf{Segmentation}: The intermediate output that the TLM generates is structured so that the system can infer from it an accurate \emph{segmentation} of the literal body that distinguishes spliced terms, i.e. those that appear in the expansion, from segments  parsed in some other way. The segmentation is all that needs to be communicated to language services downstream of the expander, e.g. editors and pretty printers, which can pass it on to the programmer using secondary notation, e.g. colors in this document. So by examining Line 9, the programmer knows that the two instances of \li{x} are spliced expressions (because they are in black), whereas the \li{R}'s must be parsed in some other way, e.g. as operators of the K language (because they are in green). This also implies that errors in spliced terms can always be reported in terms of their original location.
\item \textbf{Segment Typing}: Each spliced segment in the inferred segmentation also has  a type annotation. This, together with the context independence condition, ensures that type inference at the TLM application site can be performed (by editor services or in the programmer's mind) abstractly, i.e. by reference only to the type annotations on the spliced segments, not the full expansion.  %This information is usually not necessary to reason about typing, but it can be conveyed to the programmer upon request by the program editor if desired. %TLM definitions follow the usual scoping rules, so it is easy to ``jump to the definition'' of \li{$kq}.
\item \textbf{Capture}: Splicing is guaranteed to be capture-avoiding, so the spliced expression \li{x} must refer to the binding of \li{x} on Line 2. It cannot have captured a coincidental binding of \li{x} in the expansion. % We will say more about capture later on.
\end{enumerate}

@@

\paragraph{Paper Outline} 
% \begin{itemize}
% \item 
Sec.~\ref{sec:setsms} details expression TLMs in Reason with a more detailed case study. 
% \item 
Sec.~\ref{sec:sptsms} then introduces {pattern TLMs} and describes the special reasoning conditions in pattern position. %The hygiene condition for pattern TLMs is interesting.
% \item 
Sec.~\ref{sec:ptsms} introduces the more general {parametric TLMs}, which allow us to define literal notation at a type- or module-parameterized family of types. This section also reveals that the \li{expansions require} clause can be understood as parameterization followed immediately by partial parameter application. Having introduced the basic machinery, we continue in Sec.~\ref{sec:more-examples} with examples that further demonstrate the expressive power of this approach and various parser implementation strategies. Notably, this section gives our take on the example from  Fig.~\ref{fig:urweb} of Ur/Web-style HTML literals as well as a TLM that implements quasiquotation for Reason language terms, which is useful for implementing other TLMs. 
Sec.~\ref{sec:setsms-formally} defines a type-theoretic calculus of simple expression and pattern TLMs and formally establishes the reasoning principles implied above in their simplest form. 
Sec.~\ref{sec:ptlms-formally} adds type functions and an ML-style module system to this calculus, and gives a more general variant of the reasoning principles theorem. 
Sec.~\ref{sec:implementation} provides a brief overview of how we are implementing TLMs  without modifying OCaml's type system. 
% \item 
Sec.~\ref{sec:existing-approaches} compares TLMs to related work, guided by the rubric of  reasoning principles just discussed. 
% \item 
Sec.~\ref{sec:discussion} concludes with a summary of the contributions of this paper and a discussion of limitations and future work.
Certain technical details, proofs and extensions to the presented calculi are available in the supplement. % Also available is an additional series of case studies: notation for an encoding of regular expressions as an abstract data type, implemented using TLMs for quasiquotation and grammar-based parser generators. %are also available in the supplement. Sec.~\ref{sec:static-eval} considers the topic of term evaluation during the  expansion phase, i.e. \emph{static evaluation}, in  more detail. Sec \ref{sec:static-eval} also gives examples of TLMs that are useful for defining other TLMs, e.g. TLMs that implement parser generators and quasiquotation.

@@

Consider the recursive datatype \li{Regex.t} defined by Fig.~\ref{fig:Regex-module-def}, which encodes {regular expressions} (regexes) into Reason \cite{Thompson:1968:PTR:363347.363387}. Regexes are  common in, for example, bioinformatics, where they are used to express patterns in DNA sequences. We might construct a regex that matches the strings \li{"A"}, \li{"T"}, \li{"G"} or \li{"C"}, which represent the four bases in DNA, as follows:
\begin{lstlisting}[numbers=none]
  let any_base = Regex.(Or(Str "A", Or(Str "T", Or(Str "G", Or(Str "C")))))
\end{lstlisting}
In Reason, the notation \li{Regex.(}$e$\li{)} locally opens the module \li{Regex} within the expression $e$, so we did not need to qualify each application of \li{Regex.Or} and \li{Regex.Str} above. Even with the aid of this shorthand, however, constructing regexes in this way is syntactically costly. Instead, we would like to have the option to use the common POSIX-style notation  \cite{STD95954} when constructing values of type \li{Regex.t}, including values constructed compositionally from other regexes and strings. 
We solve this problem in Fig. \ref{fig:regex-tlm-def} by defining a TLM named \li{$regex} (pronounced ``lit regex'') that supports POSIX-style regex notation extended with splice forms for regexes and strings. Fig. \ref{fig:first-tlm-example} shows three examples of \li{$regex} being applied. 
% One approach would be to simply define a function \li{parse_rx : string -> Regex.t} that parses a string at run-time to produce a regex. This approach works for simple expressions like the one above -- we can write \li{parse_rx "A|T|G|C"}. However, this approach breaks down when one needs to construct regexes compositionally. For example, the DNA pattern that the BisI restriction enzyme recognizes is \li{AT}$x$\li{GC}, where $x$ is any of the four DNA bases. However, one cannot pass \li{any_base} to the string concatenation operator, because \li{any_base} is already of type \li{Regex.t}. These kinds of situations cause programmers to engage in \emph{stringly-typed programming}, as discussed in Sec.~\ref{sec:intro} -- they simply use strings to encode regexes throughout their application, rather than structured encodings like \li{Regex.t}. This can be catastrophic if the programmer is not extremely careful with string concatenation because strings that happen to contain special characters recognized by the POSIX notation can be misinterpreted because they are not distinguished in any way from strings that intentionally use POSIX notation. When some of these strings come from untrusted sources, e.g. web forms as is common in bioinformatics applications, the result can be a breach in privacy.

% Moreover, there already exists a broadly adopted shorthand notational standard -- the POSIX standard for regular expressions \cite{STD95954}. For the purposes of this section, we will stick to the subset of the POSIX standard that applies to the encoding in Fig.~\ref{fig:Regex-module-def}, but there is nothing that would fundamentally prevent us from implementing the full POSIX standard using TLMs.
\begin{figure}[t]
\begin{subfigure}[t]{0.45\textwidth}
\vspace{-1px}
\begin{lstlisting}[mathescape=~]
  module Regex = {
    type t = AnyChar 
           | Str(string)
           | Seq(t, t) 
           | Or(t, t) 
           | Star(t);
  };
\end{lstlisting}
\vspace{-5px}
\caption{The \li{Regex} module, which defines the recursive datatype \li{Regex.t}.}
\label{fig:Regex-module-def}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.53\textwidth}
\vspace{-1px}
\begin{lstlisting}[mathescape=|]
  module RegexNotation {
    notation $regex at Regex.t
      lexer   RegexLexer
      parser  RegexParser.start
      expansions require
        module Regex as Regex;
  };
\end{lstlisting}
\vspace{-5px}
\caption{The definition of the \li{$regex} TLM. Figure \ref{fig:lexer-and-parser} defines \li{RegexLexer} and \li{RegexParser}.}
\label{fig:regex-tlm-def}
\end{subfigure}

\begin{subfigure}[t]{\textwidth}
\vspace{4px}
\begin{lstlisting}
  open RegexNotation;
  module DNA = { let any_base = $regex `(SURLA|T|G|CEURL)`; };
  let bisA = $regex `(SURLGC$(EURLDNA.any_baseSURL)GCEURL)`;
  let restriction_template(gene) => 
    $regex `(SURL$(EURLbisASURL)$(EURLDNA.any_baseSURL)*$$(EURLgeneSURL)$(EURLDNA.any_baseSURL)*$(EURLbisASURL)EURL)`;
\end{lstlisting}
\vspace{-4px}
\caption{Examples of the \li{$regex} TLM being applied in a bioinformatics application. %In each case, literal body, between backticks, is initially left unparsed according to the language's context-free syntax. %The applied TLM determines a segmentation and expansion during the typed expansion phase, which generalizes the usual typing phase.
}
\label{fig:first-tlm-example}
\end{subfigure}
\vspace{0px}
\caption{Case Study: POSIX-style regex literal notation, with support for string and regex splicing.}
\vspace{-6px}
\label{fig:regex-case-study}
\end{figure}

\vspace{-3px}
\subsection{Client Perspective}
Let us start from the perspective of a client programmer examining Fig. \ref{fig:regex-case-study} but holding the underlying expansion, as well as the details of the lexer and parser, \li{RegexLexer} and \li{RegexParser}, abstract. We will return to describe the lexer and parser from the provider's perspective in Sec. \ref{sec:provider-perspective}.
